A Type-Theoretical Wide-Coverage Computational Grammar for Swedish

### 2012-05-19 tsd/TSD2012/paper.tex
comments from jury

\begin{abstract}
The work describes a wide-coverage computational grammar for Swedish. 
It is developed using GF (Grammatical Framework), a %dependently-typed 
functional language specialized for grammar programming. We trained
and evaluated the grammar by using Talbanken, one of the largest treebanks for 
Swedish. As a result 65$\%$ of the Talbanken trees were translated
into the GF format in the training stage and 76\% of the noun phrases
was parsed during the evaluation. Moreover, we obtained a language
model for Swedish which we use for disambiguation when parsing free text
with the grammar. 
\keywords{grammar-formalism, computational grammar, treebank}
\end{abstract}

@@

\textbf{Resource grammar}
%Chapter 2.1.2
The GF package provides an useful resource library \cite{gf-resource}, covering the
fundamental morphology and syntax of more than 20 languages.
There is also a small test lexicon included, containing a few hundred common
words.
%Since the languages all share the same abstract syntax translation between any given
%pair is possible, which is valuable
%when implementing multilingual applications.
The grammars describe how to construct phrases and sentences and how to
decline words. They cover the 
word order, agreement, tense, basic conjunction etc.
Due to the syntactic similarities between the Scandinavian languages, much of
the implementation for Swedish is shared with Norwegian and Danish. The modules
that concern the lexical aspects are separate, while 85\% of the syntax
description is shared.  There are about 80 functions, which describe the rules
for building phrases and clauses, such as functions for predication and
complementation:\\
\small \texttt{
\indent PredVP ~~~~~ : NP -> VP -> Cl ; -- Predication\\
\indent ComplVPSlash : VPSlash  -> VP ;~ -- Complementation}\\
\normalsize
%\small
%\begin{verbatim}
%PredVP       : NP -> VP -> Cl ; -- Predication
%ComplVPSlash : VPSlash  -> VP ; -- Complementation
%\end{verbatim}
%\normalsize
%The analysis preformed by GF is driven by the parts of speech, which are combined into
%parts of sentences.
In addition to the core resource grammars, which is shared with all other
languages implemented in the library, there is also an extra module
%, simple called the \verb|Extra| module,
where language specific constructions may be added.\\
%The functions given here do not have to be translatable
%to all other language, but are meant to cover language specific constructions.

@@

Our solution introduces the use of dependent types.
We make a difference between subjects and objects
%but not by giving them entirely different types, but
by letting the type \verb-NP- depend on an argument, 
which may either be \verb-Subject- or \verb-Object-.\\
\small \texttt{
\indent PredVP  ~~~~~ : NP Subject -> VP        -> Cl ; ~~~ -- predication\\
\indent ComplVPSlash : VPSlash    -> NP Object -> VP ; -- complementation\\
\indent PrepNP       : (a: NPType) -> Prep -> NP a -> Adv a ; -- adverbial phrase \\
}
\normalsize
We hence combine the-part-of-speech driven analysis normally performed by
GF with a part-of-sentence analysis, where the dependent types give the 
information we were missing.
%The dependent types in GF have been used for logical reasoning before, but
This is the first example in GF making use of the dependent types for 
describing pure syntax of a natural language. The approach could be extended
to describing similar relations
%between noun phrases and antecedent 
in other languages.\\
%having could be extended to use the type system in similar ways.

\noindent \textbf{Overgeneration}
Another aspect of the grammar implementation was the avoidance of overgeneration. As
the Swedish resource grammar had not been used for larger projects, many examples of
overgeneration were present that did not cause problems when working with small
lexicons and controlled language, but that caused explosions of ambiguities when 
tested on Talbanken. For the function \verb-GenNP-, used for genitive, we managed
to decrease the number of parse trees for short, yet problematic sentences to 10\%.

@@

Talbanken contains very valuable information about phrase structures and word 
usage.
One part of this project has focused on translating trees from Talbanken to GF
by constructing a mapping, which automatically transforms trees in the
Talbanken format to GF abstract syntax trees. We hence get a comparison 
\begin{figure}[!h]
\centering
\subfloat[Talbanken tree]
{\includegraphics[width=60mm]{TBTree_cropped.pdf}}
\subfloat[Translation to GF abstract tree]
{\includegraphics[width=60mm]{latt2.pdf}}
\caption{The sentence \emph{Det är inte lätt} (``It is not easy").}
\label{fig:gftree1}
\end{figure}
between the two annotations and at the same time we extract a GF treebank.
Nodes that fail to be translated are represented by a meta variable,
annotated \verb-?-.  In figure \ref{fig:gftree1}, the word \emph{lätt} (``easy")
was unknown and therefore represented by a meta variable in the 
translated tree. Meta variables are also used for connecting 
sister nodes that can be translated on their own, but not joined into one tree.

Talbanken05 uses three kinds of tags: categories, edge labels and POS-tags. 
While the POS-tags are reserved for words, the categories give grammatical information:
 \verb|S|, \verb|NP| and \verb|VP|.
The edge labels give the part of sentence: \verb|SS| for subject, 
\verb|OO| for object etc. 
%
%\begin{tabular}{lll}
%\textbf{Tag} & \textbf{Example} & \textbf{Explanation} \\
%\hline
%\textbf{SV} & \emph{ska} (`shall') & The verb \emph{shall}\\
%\textbf{NNDDHSGG} & \emph{familjemedlemmarnas} & Noun, definite, compound (person),\\
%&(`the family members' ') &genitive \\
%\textbf{XX} & & Unclassifiable part-of-speech\\
%\end{tabular}\\
The mapping could in some cases be performed tag-by-tag, but annotational
differences complicated the translation.
One example is valency information, which is given implicitly by the complements
of a word in Talbanken. If a verb is followed by an object, \verb-OO- containing an
\verb-S-, we can conclude that this is a sentence-complement verb.
In GF, the valency information is given explicitly for each entry in the lexicon.
A sentence-complement verb has the type \verb-VS- and the 
function \verb-ComplVS-  must be used for combining the verb and  a 
sentence into a complete verb phrase.
%Another annotational difference that turned out to be problematic for the
%translation can be illustrated by analysis of the generic pronoun \emph{`man'}.
%In Talbanken, \emph{`man'} is simply a personal pronoun, 
%in GF it is represented by using the function \verb|GenericCl| which turns a
%\verb-VP- into a clause and has no subject visible in the abstract tree. 
%When translating a sentence like this, it is thus not possible to simply glue the
%subject to the verb with \verb|PredVP|. For each similar GF construction, an extra rule
%would be needed to get full coverage.

@@

\section{Chunk Parsing}
\label{sec:chunk}
%New section - insights about latest work
The grammar, which is hand written, cannot be expected to cover all
possible expressions of the language.
Hence, we combine it with statistical means to get an even better coverage. 
%The parser uses named entity recognition and a compounding analysis.
The parser should give as much output as possible 
%Further we need to be able to give output
even when encountering unknown words and grammatical constructions,
idioms and ellipses.
Our approach uses of the rich annotation in Talbanken -- we perform chunk parsing
relying on the tags given in the treebank. 
Consequentially, whenever a sentence is not recognized by the grammar, it is 
parsed chunk by chunk and the parse trees for each chunk are returned.
Noun, adverbial and adjectival phrases are considered.
Due to the differences in the annotation between Talbanken and
GF mentioned is Section \ref{sec:mapping}, verbs are not considered
as parsable chunks. %The GF parser depend on requires verb particles and prepositions to give
%a correct output, and which are usually not in the same Talbanken chunk. 
Therefore, our focus is on whole sentences and noun phrases. At the first stage,
we try to parse the basic sentence structure, and therefore allow the parser
to treat complicated chunks as dummy words. This way, we don not lose the 
high-level sentence analysis given by GF.
Subsequently, we try to separately parse the chunks,
until as much of the sentence as possible is covered.

@@

\section{Evaluation}

@@

\section{Conclusion}
%From 6.1 - the discussion part + something about the latest work
Our work resulted in a wide-coverage grammar for Swedish which --
combined with statistical resources -- can be used for parsing 
open-domain text. The grammar covers a large number of syntactic 
phenomena and uses an extensive morphological lexicon. To
our knowledge, it is the most comprehensive computational grammar for
Swedish and the most complex GF grammar to date.
Besides parsing, the grammar may well be used for language generation.
All parts of the project are open-source and the grammar and other parts
are modular and could be reused and further developed.

>>>>

\begin{abstract}
%\textbf{typos, omissions, grammar, proof read!}
The work describes a wide-coverage computational grammar for Swedish. 
It is developed using GF (Grammatical Framework), a %dependently-typed 
functional language specialized for grammar programming. We trained
and evaluated the grammar by using Talbanken, one of the largest treebanks for 
Swedish. As a result 65$\%$ of the Talbanken trees were translated
into the GF format in the training stage and 76\% of the noun phrases
was parsed during the evaluation. Moreover, we obtained a language
model for Swedish which we use for disambiguation when parsing free text
with the grammar. 
\keywords{grammar-formalism, computational grammar, treebank}
\end{abstract}

@@

\textbf{Resource grammar}
%Chapter 2.1.2
The GF package provides an useful resource library \cite{gf-resource}, covering the
fundamental morphology and syntax of more than 20 languages.
There is also a small test lexicon included, containing a few hundred common
words.
%Since the languages all share the same abstract syntax translation between any given
%pair is possible, which is valuable
%when implementing multilingual applications.
The grammars describe how to construct phrases and sentences and how to
decline words. They cover the 
word order, agreement, tense, basic conjunction, etc.
Due to the syntactic similarities between the Scandinavian languages, much of
the implementation for Swedish is shared with Norwegian and Danish. The modules
that concern the lexical aspects are separate, while 85\% of the syntax
description is shared.  There are about 80 functions, which describe the rules
for building phrases and clauses, such as functions for predication and
complementation:\\
%\textbf{explain notation, why ComplVPSlash?}\\
\small \texttt{
\indent PredVP ~~~ : NP -> VP -> Cl ;~~~~~~ -- Predication\\
\indent ComplSlash : VPSlash  -> NP -> VP ;~ -- Complementation}\\
\normalsize
Here the types for the functions are given -- \texttt{PredVP} will return a clause when 
given its input arguments: a noun phrase and a verb phrase.
%\small
%\begin{verbatim}
%PredVP       : NP -> VP -> Cl ; -- Predication
%ComplVPSlash : VPSlash  -> VP ; -- Complementation
%\end{verbatim}
%\normalsize
%The analysis preformed by GF is driven by the parts of speech, which are combined into
%parts of sentences.
In addition to the core resource grammars, which is shared with all other
languages implemented in the library, there is also an extra module
%, simple called the \verb|Extra| module,
where language specific constructions may be added.\\
%The functions given here do not have to be translatable
%to all other language, but are meant to cover language specific constructions.

@@

Our solution introduces the use of dependent types.
We make a difference between subjects and objects
%but not by giving them entirely different types, but
by letting the type \verb-NP- depend on an argument, 
which may either be \verb-Subject- or \verb-Object-.\\
\small \texttt{
\indent PredVP  ~~~ : NP Subject -> VP        -> Cl ; ~~~ -- predication\\
\indent ComplSlash : VPSlash    -> NP Object -> VP ; -- complementation\\
\indent PrepNP       : (a: NPType) -> Prep -> NP a -> Adv a ; -- adverbial phrase \\
}
\normalsize
We hence combine the-part-of-speech driven analysis normally performed by
GF with a part-of-sentence analysis, where the dependent types give the 
information we were missing.
%The dependent types in GF have been used for logical reasoning before, but
This is the first example in GF making use of the dependent types for 
describing pure syntax of a natural language. The approach could be extended
to describing similar relations
%between noun phrases and antecedent 
in other languages.\\
%having could be extended to use the type system in similar ways.

\noindent \textbf{Overgeneration}
%\textbf{how many at first, how about other sentences?}
Another aspect of the grammar implementation was the avoidance of overgeneration. As
the Swedish resource grammar had not been used for larger projects, many examples of
overgeneration were present that did not cause problems when working with small
lexicons and controlled language, but that caused explosions of ambiguities when 
tested on Talbanken. For the function \verb-GenNP-, used for genitive, we managed
to decrease the number of parse trees for short, yet problematic sentences to 10\%.

@@

%\textbf{About the underlying probalistic model of GF}
Talbanken contains very valuable information about phrase structures and word 
usage.
One part of this project has focused on translating trees from Talbanken to GF
by constructing a mapping, which automatically transforms trees in the
Talbanken format to GF abstract syntax trees. We hence get a comparison 
\begin{figure}[!h]
\centering
\subfloat[Talbanken tree]
{\includegraphics[width=60mm]{TBTree_cropped.pdf}}
\subfloat[Translation to GF abstract tree]
{\includegraphics[width=60mm]{latt2.pdf}}
\caption{The sentence \emph{Det är inte lätt} (``It is not easy").}
\label{fig:gftree1}
\end{figure}
between the two annotations and at the same time we extract a GF treebank.
Nodes that fail to be translated are represented by a meta variable,
annotated \verb-?-.  In figure \ref{fig:gftree1}, the word \emph{lätt} (``easy")
was unknown and therefore represented by a meta variable in the 
translated tree. Meta variables are also used for connecting 
sister nodes that can be translated on their own, but not joined into one tree.

Talbanken05 uses three kinds of tags: categories, edge labels and POS-tags. 
While the POS-tags are reserved for words, the categories give grammatical information:
 \verb|S|, \verb|NP| and \verb|VP|.
The edge labels give the part of sentence: \verb|SS| for subject, 
\verb|OO| for object, etc. 
%
%\begin{tabular}{lll}
%\textbf{Tag} & \textbf{Example} & \textbf{Explanation} \\
%\hline
%\textbf{SV} & \emph{ska} (`shall') & The verb \emph{shall}\\
%\textbf{NNDDHSGG} & \emph{familjemedlemmarnas} & Noun, definite, compound (person),\\
%&(`the family members' ') &genitive \\
%\textbf{XX} & & Unclassifiable part-of-speech\\
%\end{tabular}\\
The mapping could in some cases be performed tag-by-tag, but annotational
differences complicated the translation.
One example is valency information, which is given implicitly by the complements
of a word in Talbanken. If a verb is followed by an object, \verb-OO- containing an
\verb-S-, we can conclude that this is a sentence-complement verb.
In GF, the valency information is given explicitly for each entry in the lexicon.
A sentence-complement verb has the type \verb-VS- and the 
function \verb-ComplVS-  must be used for combining the verb and  a 
sentence into a complete verb phrase.
%Another annotational difference that turned out to be problematic for the
%translation can be illustrated by analysis of the generic pronoun \emph{`man'}.
%In Talbanken, \emph{`man'} is simply a personal pronoun, 
%in GF it is represented by using the function \verb|GenericCl| which turns a
%\verb-VP- into a clause and has no subject visible in the abstract tree. 
%When translating a sentence like this, it is thus not possible to simply glue the
%subject to the verb with \verb|PredVP|. For each similar GF construction, an extra rule
%would be needed to get full coverage.

@@

\section{Chunk Parsing}
\label{sec:chunk}
%\textbf{what does eval in terms of metas mean? is gf not necessary when chunking}
%New section - insights about latest work
The grammar, which is hand written, cannot be expected to cover all
possible expressions of the language.
Hence, we combine it with statistical means to get an even better coverage. 
%The parser uses named entity recognition and a compounding analysis.
The parser should give as much output as possible 
%Further we need to be able to give output
even when encountering unknown words and grammatical constructions,
idioms and ellipses.
Our approach uses of the rich annotation in Talbanken -- we perform chunk parsing
relying on the tags given in the treebank. 
Consequentially, whenever a sentence is not recognized by the grammar, it is 
parsed chunk by chunk. Each chunk is analysed by the grammar-driven parser, and
resulting trees are returned.
Noun, adverbial and adjectival phrases are considered.
Due to the differences in the annotation between Talbanken and
GF mentioned is Section \ref{sec:mapping}, verbs are not considered
as parsable chunks. %The GF parser depend on requires verb particles and prepositions to give
%a correct output, and which are usually not in the same Talbanken chunk. 
Therefore, our focus is on whole sentences and noun phrases. At the first stage,
we try to parse the basic sentence structure, and therefore allow the parser
to treat complicated chunks as dummy words. This way, we don not lose the 
high-level sentence analysis given by GF.
Subsequently, we try to separately parse the chunks,
until as much of the sentence as possible is covered.

@@

\section{Evaluation}
%\textbf{limited, how many rules? 45 functions added, plus
%categories, updates, and tense}

@@

\section{Conclusion}
%\textbf{only minimal success? not ripe enough?}
%From 6.1 - the discussion part + something about the latest work
Our work resulted in a wide-coverage grammar for Swedish which --
combined with statistical resources -- can be used for parsing 
open-domain text. The grammar covers a large number of syntactic 
phenomena and uses an extensive morphological lexicon. To
our knowledge, it is the most comprehensive computational grammar for
Swedish and the most complex GF grammar to date.
%\textbf{based on what?}
Besides parsing, the grammar may well be used for language generation.
All parts of the project are open-source and the grammar and other parts
are modular and could be reused and further developed.
